---
title: "ASV_scatter"
author: "Harald Brolin"
date: "September 5, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = TRUE,
	warning = TRUE
)
```

```{r message=FALSE, warning=FALSE}
library("phyloseq"); packageVersion("phyloseq")
library("ggplot2"); packageVersion("ggplot2")
library("tidyverse")
library("gridExtra")
library("Hmisc")
library("knitr")
library("EnvStats")

df <- readRDS("../phyloseq_v2.rds")
```
## Analysis of amplicon sequence variants (ASV) 
This data has not been rarefied nor has any singletons (or any ASVs for that matter) been removed.
First we want to investigate the ASVs

```{r all Asv}

# Create df with ASV-stats
df_otu <- data.frame(row.names = 1:nrow(df@otu_table), # rownames as numbers
                   asv = df@otu_table %>% rownames(), # asvs
                   nr_samples = apply(df@otu_table, 2, function(x) (x != 0)) %>% rowSums(), # Count number of sites in which the OTU occurs in
                   abundence = df@otu_table %>% rowSums(), # Total number of reads
                   variance = apply(df@otu_table, 1, function(x) (var(x))))

re_abundence <- apply(df@otu_table, 2, function(x) (x / sum(x) )) %>% as.data.frame() # Calulate fraction of reads for all samples
df_otu$re_abundence <- apply(re_abundence, 1, function(x) (mean(x))) # mean of fraction
df_otu$re_variance <- apply(re_abundence, 1, function(x) (var(x))) # variance of fraction 

# Count the number of sites each ASV exists in
count_number_of_sites <- function(loc) {
  df_loc <- prune_samples(get_variable(df, "location") == loc, df) # Filter based on site
  asv_loc <- # Creates the df with TRUE / FALSE for each ASV dependent on prescence in location  
    data.frame(row.names = 1:nrow(df_loc@otu_table),
                count  = (apply(df_loc@otu_table, 2, function(x) (x != 0)) %>% rowSums()) > 0) # For each sample and each row, does the ASV have a count larger than 0
  names(asv_loc) <- paste(loc , "exists", sep = "_") # Create column name, Stockholm_exists
  return(asv_loc)
}

# Apply the function
df_number_of_sites <- # Apply function to all locations, column bind a final df
  do.call(cbind, 
          lapply(df@sam_data$location %>%
                   levels, count_number_of_sites))

df_number_of_sites$asv <- df@otu_table %>% rownames() # Get rownames, they are in the same order as df@otu_table
df_otu$number_of_sites <- df_number_of_sites[,1:6] %>% rowSums() %>% as.factor()# Calculate the the number of lakes of which the ASV occurs

p1 <- ggplot(df_otu , aes(x = nr_samples, y = re_abundence)) + geom_point(aes(color = number_of_sites)) + scale_y_log10() + ggtitle("Scatterplot of mean relative abundence and ASV occurence"); p1
ggsave(plot = p1, filename = "../../analysis_images/5_asv_scatter_sites.png")

df_otu$Phylum <- df@tax_table[,"Phylum"]
# df_otu$Class <- df@tax_table[, "Class"] 
# df_otu$Order <- df@tax_table[, "Order"]
# df_otu$Family <- df@tax_table[,"Family"]

get_top_taxa <- function(taxa){
    df_phylum <- tax_glom(df, taxrank = taxa, NArm = FALSE)
    top_phylum <- df_phylum@otu_table %>% rowSums()
    top_6 <- top_phylum %>% order(., decreasing = TRUE) %>% head(6) # Arrange taxa by reads, select top 5 (+ NA)
    return(df_phylum@tax_table[top_6, taxa] %>% as.vector)
}

taxa_colors <- get_top_taxa("Phylum")
df_otu$Phylum[which( ! (df_otu$Phylum %in% taxa_colors))] <- "Other" # The taxa not in top_8, set to other
df_otu$Phylum <- df_otu$Phylum %>% addNA()

p2 <- 
  ggplot(df_otu,aes(x = nr_samples, y = re_abundence)) +
  geom_point(aes(color = Phylum)) +
  scale_y_log10() +
  ggtitle("Scatterplot of mean relative abundence and ASV occurence"); p2

p2 <- 
  ggplot(df_otu %>% filter(nr_samples > 50),aes(x = nr_samples, y = re_abundence)) +
  geom_point(aes(color = Phylum)) +
  scale_y_log10() +
  ggtitle("Scatterplot of mean relative abundence and ASV occurence"); p2

p3 <- ggplot(df_otu , aes(x = nr_samples, y = re_abundence)) +
  geom_point(aes(color = Phylum)) +
  scale_y_log10() +
  ggtitle("Scatterplot of mean relative abundence and ASV occurence") +
  facet_wrap(~Phylum); p3

ggsave(plot = p3, filename = "../../analysis_images/5_asv_scatter_phylum.png")

ggplot(df_otu, aes(y = nr_samples, x = Phylum)) +
  geom_boxplot(aes(color = Phylum), show.legend = FALSE) +
  scale_y_log10() + 
  coord_flip() + 
  facet_wrap(~number_of_sites) 

# df_bar <- 
#   df_otu %>% 
#   group_by(Phylum) %>%
#   summarise(phylum_cnt = sum(abundence)) %>% 
#   mutate(taxa_sites = count / sum (count))
  

ggplot(df_otu, aes(number_of_sites)) + geom_bar(aes(fill = Phylum)) + scale_y_log10()

ggplot(df_otu , aes(x = nr_samples, y = re_abundence)) +
  geom_point(aes(color = log10(re_variance))) +
  scale_y_log10() +
  ggtitle("Scatterplot of mean relative abundence and ASV occurence")

```
#### Conclusion
Similar patten can be seen here as in the previous data (filter 130), although we have 35,000 more ASVs. It is also interesting to note that the spread of varance is much smaller, I do not know why this is the case (maybe the difference lies in the use of baseMean, a measurment which is not capped as relative_abundence at 1. It might also be due to the amount of retained reads have almost half the amount of reads comapired to the previous run)
