---
title: "Sjobo table"
author: "Harald Brolin"
date: "8 augusti 2018"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```


```{r cars}
library("phyloseq")
library("ggplot2")
library("tidyverse")
library("gridExtra")
library("knitr")

df <- readRDS("../../phyloseq.rds") # Load phyloseq-object
```

# Sample select and phyloseq-data agglomeration 
# (Finns risk att OTU:er som bara finns i ett prov kommer med, troligtvis inte på mean men ev på differnce och variance)



```{r pressure, echo=TRUE}
sequences <- df@otu_table %>% rownames()
taxa_names(df) <- 1:(df@otu_table %>% nrow)
  
samples <- c("Sjobo140306", "Sjobo140307kl0900", "Sjobo140307kl1200",
             "Sjobo140318","Sjobo140404")

df@sam_data$SampleName <- rownames(df@sam_data) # Add the sample ID as a column
df <- subset_samples(df, SampleName %in% samples) # Choose samples of interest
df <- prune_taxa(taxa_sums(df) > 4, df) # Remove OTU:s not included in theese samples
df_trans <- transform_sample_counts(df, function(x)  x/sum(x) * 100 ) # Transform the taxa_table into percent of total reads (%)

df_trans <- tax_glom(df_trans, 'Phylum',NArm=F)
```

The **mean** of the downpour samples.

The **variance** between the downpour samples and the "normal" samples.

The **difference** between the downpour samples and the sample taken befor the downpour.

```{r echo=TRUE}
# Convert the otu_table to a tibble
df <- df_trans@otu_table %>% 
      as.data.frame() %>% 
      rownames_to_column() %>% 
      as.tibble()

downpour_samples <- c("Sjobo140307kl0900", "Sjobo140307kl1200")

# Create column of percent means for the downpour samples
df$mean_downpour_samples <- df %>% 
                            select(downpour_samples) %>% rowMeans()

# Get the mean for an OTU, using the mean for the downpour sampes
df$mean <- df %>% 
           select(Sjobo140306, mean_downpour_samples, Sjobo140318, Sjobo140404) %>%
           rowMeans()

# Get the variance in percentage, using mean for the downpour samples 
df$variance <- df %>%
               select(Sjobo140306, mean_downpour_samples, Sjobo140318, Sjobo140404) %>% 
               apply(., 1, var)

# Get the difference in percentage betweeen befor the downpour and the mean of the downpour samples
diff_before <- df %>% 
               mutate(difference = mean_downpour_samples - Sjobo140306) %>%
               select(difference)

# To associate the taxonomic information with the df
df <- cbind(df, diff_before) %>% as.data.frame %>% as.tibble()

# To add the taxonomic classifications
df <- cbind(df,
            df_trans@tax_table[,1:2] %>% 
            as.data.frame() %>% 
            as.tibble()) %>% as.tibble()

names(df)[1] <- "asv_id"

```




## Table over the main Phylum, here ordered by greatest **difference** in percentage (%) (before and after)

```{r}
# A negative value indicated a decrease in abundence while a positive value indicates an increase
diff_10 <- df %>% arrange(desc(abs(difference))) %>% select(1,2,4,5,3,6,8,9,10,11,12)
kable(diff_10[1:10,],
      digits = 3,
      align = "c",
      caption = "Arranged by the greatest difference in percentage of reads")
```




## Table over the main Phylum, here ordered by the size of the **mean** percentage (%)

```{r}
mean_10 <- df %>% arrange(desc(mean)) %>% select(1,2,4,5,3,6,8,9,10,11,12)
kable(mean_10[1:10,], 
      digits = 3,
      align = "c",
      caption = "Arranged by the greatest mean percentage of reads")
```




## Table over the main Phylum, here ordered by the **variance** of the percentage (%) 

```{r}
var_10 <- df %>% arrange(desc(variance)) %>% select(1,2,4,5,3,6,8,9,10,11,12)
kable(var_10[1:10,],
      digits = 3,
      align = "c",
      caption = "Arranged by the greatest variance of percentage of reads")
```


# From the table ordering the greatest **difference**, we can se that the largest difference is in the "Unclassified" group. 
This group can be the agglomeration of several "Unclassified" OTUs (though the barplot indicates that group mainly consists of one OTU) 
and thus further investigation is required.

In this caseI will only look at "Sjobo140307kl0900" since both downpour samples exhibiths similar "OTU-fingerprints".

```{r echo=TRUE, message=FALSE}
df <- readRDS("../../phyloseq.rds") # Load phyloseq-object

sequences <- df@otu_table %>% rownames()
taxa_names(df) <- 1:(df@otu_table %>% nrow)

samples <- c("Sjobo140307kl0900","Sjobo140306")
df@sam_data$SampleName <- rownames(df@sam_data) # Add the sample ID as a column
df <- subset_samples(df, SampleName %in% samples) # Choose samples of interest
df <- prune_taxa(taxa_sums(df) > 1, df) # Remove OTU:s not included in theese samples

unclassified_otus <- df@tax_table[which(is.na(df@tax_table[,2])), 2] %>% rownames()  # Select only "Unclassified", denoted in this dataframe as ""
df_trans <- transform_sample_counts(df, function(x)  x/sum(x) * 100 )  # Convert into percent
df_unclassified <- df_trans@otu_table[unclassified_otus,] %>% as.data.frame() %>% rownames_to_column()

df_unclassified$difference <- df_unclassified[,3] - df_unclassified[,2]  # Subtract the OTU-percent to get the difference
df_print <- df_unclassified %>%
            as.tibble() %>% 
            arrange(desc(abs(difference)))

names(df_print)[1] <- "asv_id" 


# Do the same for all ASV:s
df_all <- df_trans@otu_table %>% as.data.frame() %>% rownames_to_column()

df_all$difference <- df_all[,3] - df_all[,2]  # Subtract the OTU-percent to get the difference

df_all$mean_relative_difference <- df_all$difference / rowMeans(df_all[,2:3])


df_print <- df_all %>%
            as.tibble() %>% 
            arrange(desc(abs(difference)))

kable(df_print[1:10, ],
      digits = 3,
      align = "c",
      caption = "Arranged by the greatest variance of percentage of reads")

kable(df@tax_table[df_print$rowname[1:10],], align = "c")

df_print <- df_print %>% filter(Sjobo140306 > 0 & Sjobo140307kl0900 > 0) %>% arrange(desc(abs(mean_relative_difference)))

kable(df_print[1:10, ],
      digits = 5,
      align = "c",
      caption = "Arranged by the greatest variance of percentage of reads")

sequences[c(24151, 3900, 3926)]

```


When filtering OTUs included in both "Sjobo140307kl0900" and "Sjobo140306" we run the risk of removing highly abundant OTUs included only in one of the two samples. 
This case was investigated and in the file "data/unclassified_otus_sjobo_downpour.rds" we can see that the top 3 otus are the same (i.e. they exist in both samples).
Since the OTUs are unclassified, we can do a ncbi-BLAST.

